A thermodynamic view of dusty protoplanetary disks

### 2017-05-09 paper.tex
misc edits to the main text overall

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading introduce buoyancy forces that generally 
  stabilizes the vertical shear 
  instability that arises from a radial temperature gradient. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes. %We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

@@

Another approach is to model the dust population as a continous, pressureless
fluid 
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of, and adapt, already-developed numerical methods/codes for
simulating pure gas dynamics.

@@

The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for two-fluid models and 
following Lagragian particles; but is straight forward to simulate in the 
one-fluid formulation \citep{price15}.  The single fluid formalism
also has the important advantage that it is 
amenable to standard analytical methods \citep{youdin05a,jacquet11}.  
This is the necessary approach to gain physical insight.

@@

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a, 
  jacquet11,laibe14}. %This eliminates the evolutionary equation for
%$\Delta \bm{v}$. 
 We also consider irradiated protoplanetary disks 
in which the temperature is set externally
\citep{chiang97,stam08}. This allows us to adopt a locally isothermal
equation of state, $P = c_s^2\rhog$, where $c_s$ is a prescribed
sound-speed profile. This physically represents short cooling times
\citep{lin15}.

Under these approximations, we show that the equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is analogous to the advection
of entropy following an adiabatic fluid. We will find that the   
effective entropy of isothermal dusty gas scales with its gas
fraction. Finite dust-gas drag act as effective heating/cooling: the relative
drift between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. %We explicitly demonstrate this for
%perfectly coupled dust in strictly and locally isothermal disks. 
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We put forward a thermodynamical interpretation of
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. In \S\ref{key_idea} we first
describe, without derivation, the analogy between dusty gas and    
ideal hydrodynamics. We then discuss in 
\S\ref{grow_osc} a general result of fluid dynamics that
underpins the bulk of our study: intabilities due to 
pressure-density phase lags. This means that dusty-gas phenomena, such
as SI, have thermodynamical interpretations.

@@

\section{Dust-free modeling of dusty protoplanetary disks}\label{key_idea}
{\bf is this section needed?} 
The key concept we introduce in this work is the description
of dusty protoplanetary disks \emph{without} having dust as an
explicit variable. This is best illustrated by writing down our model  
equations that describe a mixture of isothermal gas and small
dust particles: 
\begin{align}  
  &\frac{\p \rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right)= 0, \label{mass0}\\ 
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} = -
  \frac{1}{\rho}\nabla  P - \nabla \Phi_*,\label{mom0}\\ 
 & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P 
  = - P \nabla\cdot\bm{v}
                + \mathcal{C}(\rho, P; \tstop)\label{en0}, 
\end{align}
where $\rho=\rhod + \rhog$ is the total density, $P$ is the pressure,
$\Phi_*$ is the stellar gravitational potential,   
and $\mathcal{C}$ represent effects of dust-gas drag, which
is paremterized by the stopping time $\tstop$.

Eq. \ref{mass0}---\ref{en0} appear identical to adiabatic gas dynamics, even  
though they physically represents a two-phase, dusty-gas. The main
difference is the term $\mathcal{C}$. Here it does not represent
physical cooling (as the system is isothermal). Instead, dust-gas drag
causes particles to drift, hence by momentum conservation a gas drift
in the opposite direction, which changes the pressure distribution.

Since the above equations are similar to ideal gas dynamics, it 
means that many of the results derived for ideal hydrodynamics can be  
generalized to dusty disks without additional effort. This includes 
the important effect of phase lags between density and pressure in
fluids, discussed next.

\section{Growing oscillations by doing work}\label{grow_osc}
A general result in fluid dynamics is that work is done whenever the
pressure and density evolution of the fluid are not in phase. If 
the average work done is positive, then oscillation amplitudes 
grow: the work done allows the system to `overshoot' beyond
the amplitude of preceeding oscillations.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of growing oscillations in a
    fluid. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the P-V plane. In the case shown, 
    dust-drag causes 
    pressure to lag behind density, which results in a clockwise path in
    the P-V plane, implying positive work done by the fluid and
    hence would lead to growing oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with the cylindrical radius
$r$ from the star \citep{chiang97}. Since $P\propto \rhog T$ for an
ideal gas, the pressure of a gas parcel varies with  
position as well. There is no reason to expect $P$ and $\rhog$ to
be in phase as a gas parcel oscillates between different radii. In
fact, we will show that this is a fundamental property of the vertical
shear instability \citep{lin15}.

@@

In order to apply this thermodynamic interpretation of dust-gas drag
instabilities, we need to develop an analogy between dusty-gas
and pure gas dynamics. We show this is possible in the limit of
strong drag and a fixed gas equation of state.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an ideal gas.   
  Motivated by this similarity, we develop a thermodynamic 
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
  the streaming instability neccessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduce buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of this thermodynamic framework in
finding dusty analogs of other hydrodynamic 
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), in simulating dusty protoplanetary disks
using purely hydrodynamic codes, as well as potential generalizations
of the dusty/adiabatic gas equivalence principle.

%We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

@@

Another approach is to model the dust population as a continous, pressureless
fluid 
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of existing numerical methods/codes for
simulating pure gas dynamics.

@@

The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for two-fluid models and 
following Lagragian particles; but is straight forward with the 
one-fluid formulation \citep{price15}.  The single fluid framework 
also has the important advantage that it is 
amenable to standard analytical methods \citep{youdin05a,jacquet11}.  
This is the necessary approach to gain physical insight.

@@

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set the dust-gas relative drift
to $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a, 
  jacquet11,laibe14}. This approximation reflects the physical result
that dust particles tend to drift towards pressure maximum.

 We also consider irradiated protoplanetary disks 
 in which the temperature is set externally
\citep{chiang97,stam08}. This allows us to adopt a locally isothermal
equation of state, $P = c_s^2\rhog$, where $c_s$ is a prescribed
sound-speed profile. This physically represents short cooling times
\citep{lin15}.

Under these two approximations, we show that the equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is analogous to the advection
of entropy following an adiabatic fluid. We show that the   
effective entropy of isothermal dusty gas scales with its gas
fraction. Finite dust-gas drag act as effective heating/cooling: the relative
drift between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into standard  
hydrodynamics, we can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations.%, regardless of how  thin the dust
%layer is. 
{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. We first present, without
derivation, the correspondence between dusty gas and     
ideal hydrodynamics in \S\ref{key_idea}. In \S\ref{grow_osc} 
we discuss a key idea in fluid dynamics, relevant to our 
discussion on  SI and VSI: instabilities associated with
pressure-density phase lags.   
%This means that dusty-gas phenomena, such 
%as SI and VSI, have thermodynamical interpretations.

@@

\section{Dust-free modeling of dusty protoplanetary disks}\label{key_idea}
{\bf is this section needed?} 
The key concept we introduce in this work is the description
of dusty protoplanetary disks \emph{without} explicit dust
equations. This is best illustrated by writing down our model   
equations for a mixture of isothermal gas and small
dust particles: 
\begin{align}  
  &\frac{\p \rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right)= 0, \label{mass0}\\ 
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} = -
  \frac{1}{\rho}\nabla  P - \nabla \Phi_*,\label{mom0}\\ 
 & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P 
  = - P \nabla\cdot\bm{v}
                + \mathcal{C}(\rho, P; \tstop)\label{en0}, 
\end{align}
where $\rho=\rhod + \rhog$ is the total density, $P$ is the pressure,
$\Phi_*$ is the stellar gravitational potential,   
and $\mathcal{C}$ represent effects of dust-gas drag, which
is paremterized by the stopping time $\tstop$ ($\mathcal{C}=0$ for
$\tstop=0$). Similar equations for dusty-gas are given by
\cite{price15}.

Eq. \ref{mass0}---\ref{en0} appear identical to adiabatic gas
dynamics, even though they physically represents a two-phase,
dusty-gas. The main difference is the term $\mathcal{C}$. Here it does
not represent physical cooling, as the system is isothermal by
assumption. Instead, dust-gas drag causes particles to drift, hence by
momentum conservation a gas drift in the opposite direction, which
changes the pressure distribution.

Since the above equations are similar to ideal gas dynamics, it 
means that many of the results derived for ideal hydrodynamics can be  
generalized to dusty disks without additional effort. This includes 
the important consequence of phase lags between density and pressure,
discussed next.

\section{Growing oscillations by doing work}\label{grow_osc}
A general result in fluid dynamics is that work is done whenever
oscillations in the pressure and density of the fluid are not in
phase. If  the average work done is positive, then oscillation
amplitudes grow: the work done allows the system to `overshoot' beyond 
the amplitude of preceeding oscillations.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/drag}\\
\includegraphics[width=\linewidth]{figures/pdv}
  \caption{Thermodynamic interpretation of growing oscillations in a
    fluid. Top: pressure and density evolution in time. 
    Bottom: oscillation cycle in the P-V plane. In the case shown, 
    dust-drag causes 
    pressure to lag behind density, which results in a clockwise path in
    the P-V plane, implying positive work done by the fluid, which 
     would increase oscillation amplitudes.  
    \label{pdv_cartoon}
  }
\end{figure}

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with the cylindrical radius
$r$ from the star \citep{chiang97}. Since $P\propto \rhog T$ for an
ideal gas, %the pressure of a gas parcel varies with  
%position as well. 
there is no reason to expect $P$ and $\rhog$ to 
be in phase as a gas parcel oscillates between different radii. In
fact, we will show that this is a fundamental property of the vertical
shear instability \citep{lin15}.

@@

In order to apply this thermodynamic interpretation of dust-gas drag
instabilities, we need to develop a formal analogy between dusty-gas
and pure gas dynamics. We show this is possible in the limit of
strong drag and a fixed gas equation of state.

### 2017-05-09 result.tex
extra intro section
misc edits to the main text overall

We now present numerical solutions for vertically stratified,
non-self-gravitating disks assuming perfectly coupled dust
($\tstop=0$) so the equilibria
defined in \S\ref{eqm} are exact steady states.

Apart from \S\ref{varHd}---\ref{vert_mixed}, we consider constant
midplane dust-to-gas ratios $\epsilon_0$ and characteristic thickness
$H_\epsilon$. Then the dust-to-gas ratio $\epsilon=\epsilon(z)$. In 
this limit any growing modes must be associated with the imposed temperature gradient
(\S\ref{limits}). We are thus examining the effect of dust-loading on
the VSI previously studied in pure gas disks 
\citep[][\citetalias{lin15} in this section]{lin15}.

The parameters for this linear problem includes $\epsilon_0$, the dust
layer thickness $\Hd(r_0)$, and the perturbation radial wavenumber
$k_x$. We choose a midplane gas density profile 
$\rho_\mathrm{g0}\propto r^{-3/2}$. The fiducial power-law 
index for the temperature profile is $q=-1$; and we set the gas disk
aspect-ratio $h_\mathrm{g}=0.05$. These values were also used by  
\citetalias{lin15}.

%Two boundary conditions are needed for the $\tstop=0$
%problem \citep[e.g.][]{lubow93}. 
We impose solid vertical boundaries so that $\delta
v_z(\pm\zmax)=0$. We solve the linearized equations as a generalized
eigenvalue problem using a pseudo-spectral code adapted  
from \citetalias{lin15}. Amplitudes are expanded in Chebyshev
polynomials up to order $N_z=512$. %We check results using
%Eq. \ref{vsi_check}.

@@

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, surface modes, which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).   
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
$\epsilon_0$. i.e. it increases with the vertical buoyancy. 
The total number of modes do not change 
significantly. This is in contrast with the effect of increasing
cooling times, for which \citetalias{lin15} find fewer unstable modes.
%until the system is eventually completely stable.

@@

The lowest frequency `fundamental' body mode is energetically dominant
because the entire disk column is perturbed \citep[cf. surface modes
  which only disturb the disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$. 
This is consistent with Fig. \ref{compare_vshear_fixHd} comparing the basic state vertical
shear and buoyancy. Notice also maxium perturbtions to the dust-to-gas ratio,
$\delta\epsilon$, shifts from disk boundaries 
to the midplane.

@@

Fig. \ref{ultra_thin} shows the maximum growth rate  as a function of
the radial temperature gradient, $q$. For all cases the vertical
shear is dominated by that due to the dust layer (see
\S\ref{vertshear}). However, we see that growth rates $s\propto
|q|$, i.e .growth rates vanish in the strictly isothermal limit.  
In particular, this holds for $\rich < 0.1$, the critical value 
for non-axisymmetric instabilities.

Axisymmetric instability here is associated with the thermal
contribution to vertical shear: as $q\to0$,  $\p_z\Omega$ becomes
entirely due to dust-settling and there is no instability ($s\to
0$). This result is independent of $\rich$, so the Richardson number
is does not characterize the axisymmetric stability of dust
layers.

@@

In \S\ref{vert_mixed_theory} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.  
To demonstate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is then
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).

@@

\subsection{Validity}
{\bf maybe incorporate into intro to this section?}
We considered $\tstop=0$ in order to perform stability analysis on an
exact steady state. When $\tstop\neq0$, a stratified dusty 
disk initialized as in \S\ref{eqm} would evolve 
%due to dust-gas relative drift. In particular, in the
%absence of turbulent/diffusive processes, dust particles 
as dust settles to the
midplane on a timescale $t_\mathrm{settle}\sim 1/\tstop\OmK^2$
\citep{takeuchi02}. However, we expect the above results to remain
valid provided the VSI growth
timescale $t_\mathrm{grow}\ll t_\mathrm{settle}$. Since 
VSI growth rates are $O(h_\mathrm{g}\OmK)$, we require 
\begin{align}
  \tstop\OmK \ll h_\mathrm{g}. 
\end{align} 
For thin PPDs with $h_\mathrm{g}\sim 0.05$, this is satisfied 
for $\tstop\OmK \ll O(10^{-2})$.

>>>>

We now present numerical solutions for vertically stratified,
non-self-gravitating disks, assuming perfectly coupled dust. 
We formally take $\tstop=0$ so the equilibria defined in \S\ref{eqm} 
are exact steady states. In reality,   
%\subsection{Validity}
%{\bf maybe incorporate into intro to this section?}
%We considered $\tstop=0$ in order to perform stability analysis on an
%exact steady state. 
%When $\tstop\neq0$, a stratified dusty 
%disk initialized as in \S\ref{eqm} would evolve 
%as 
dust settles to the midplane on a timescale $t_\mathrm{settle}\sim 
1/\tstop\OmK^2$ \citep{takeuchi02}. However, we expect the 
perfectly-coupled limit to be valid  
provided timescales of interest $t_\mathrm{grow}\ll
t_\mathrm{settle}$. For the VSI this translates to 
%Since VSI growth rates are $O(h_\mathrm{g}\OmK)$, 
%we require  
%\begin{align}
 $ \tstop\OmK \ll h_\mathrm{g}. $
%\end{align} 
For thin PPDs, this is satisfied for $\tstop\OmK \ll O(10^{-2})$.

We first consider constant
midplane dust-to-gas ratios $\epsilon_0$ and characteristic thickness
$H_\epsilon$. Then the dust-to-gas ratio $\epsilon=\epsilon(z)$. In 
this limit any growing modes must be associated with the imposed
temperature gradient (see \S\ref{limits}). We are then studying the
effect of dust-loading on the VSI previously studied in pure gas
disks \citep[][\citetalias{lin15} in this section]{lin15}. In 
\S\ref{varHd} we allow $\p_r\epsilon\neq 0$, and in 
\S\ref{vert_mixed} we consider VSI driven entirely by radial
gradients in the dust-to-gas ratio (as discussed in
\S\ref{vert_mixed_theory}).

The parameters for the linear problem includes $\epsilon_0$, the dust
layer thickness $\Hd$, and the perturbation radial wavenumber
$k_x$. We choose a midplane gas density profile 
$\rho_\mathrm{g0}\propto r^{-3/2}$. The fiducial power-law 
index for the temperature profile is $q=-1$; and we set the gas disk
aspect-ratio $h_\mathrm{g}=0.05$. These values were also used by  
\citetalias{lin15}.

%Two boundary conditions are needed for the $\tstop=0$
%problem \citep[e.g.][]{lubow93}. 
We impose solid vertical boundaries so that $\delta
v_z(\pm\zmax)=0$. We solve the linearized equations as a generalized
eigenvalue problem using a pseudo-spectral code adapted  
from \citetalias{lin15}. Amplitudes are expanded in Chebyshev
polynomials up to order $512$. %We check results using
%Eq. \ref{vsi_check}.

@@

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, surface modes, which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).   
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
dust-loading, i.e. it increases with the vertical buoyancy. 
The total number of modes do not change 
significantly. This is in contrast with the effect of increasing
cooling times in an adiabatic gas disk, for which \citetalias{lin15}
find fewer unstable modes. 
%until the system is eventually completely stable.

@@

The lowest frequency `fundamental' body mode is energetically dominant
because the entire disk column is perturbed \citep[cf. surface modes
  which only disturb the disk boundaries,][]{umurhan16c}. In Fig. \ref{vsi_dust_loading2d}
we compare the fundamental mode between the nearly 
dust-free case $\epsilon_0=10^{-3}$ an a dusty disk with
$\epsilon_0=1$. Dust-loading preferentially
stabilizes the disk atmosphere against the VSI, restricting
meridional motions to $|z|\lesssim 2\Hg$. 
This is consistent with Fig. \ref{compare_vshear_fixHd} comparing the basic state vertical
shear and buoyancy. % Notice also maxium perturbtions to the dust-to-gas ratio,
% $\delta\epsilon$, shifts from disk boundaries 
% to the midplane.

@@

Fig. \ref{ultra_thin} shows the maximum growth rate  as a function of
the radial temperature gradient, $q$. For all cases the vertical
shear is dominated by that due to the dust layer (see
\S\ref{vertshear}). However, we see that $s\propto
|q|$, i.e growth rates vanish in the strictly isothermal limit.  
In particular, this holds for $\rich < 0.1$, the critical value 
for non-axisymmetric instabilities.

Axisymmetric instability here is associated with the thermal
contribution to vertical shear: as $q\to0$,  $\p_z\Omega$ becomes
entirely due to $\p_z\epsilon$ and there is no instability ($s\to
0$). This result is independent of $\rich$, so the Richardson number
is does not characterize the axisymmetric stability of dust
layers.

@@

In \S\ref{vert_mixed_theory} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.  
To demonstate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).

@@
